Temporal Continuity and the Judgment of Actual Causation

### 2021-01-31 analysis/statistical_analyses.Rmd
aurelien edits

# Experiment 1 data frame 
d1 <- read_csv(dpath_exp1) #%>%

@@

```{r exp1stats, include=TRUE, fig.height=4, fig.width=6}

d1_stats1 <- d1 %>% 
  mutate(ri = factor(case_when(response=="RC_IC" ~  NA_character_,
                        response=="RC" ~     "root",
                        response=="RD_ID" ~  NA_character_,
                        response=="RD" ~     "root",
                        response=="IC" ~     "immediate",
                        response=="ID" ~     "immediate",
                        TRUE ~ NA_character_ 
                        )),
         cd = factor(case_when(response=="RC_IC" ~  "continuous",
                        response=="RC" ~     "continuous",
                        response=="RD_ID" ~  "delayed",
                        response=="RD" ~     "delayed",
                        response=="IC" ~     "continuous",
                        response=="ID" ~     "delayed",
                        TRUE ~ NA_character_ 
                        )),
         #lengthc = sdif_coding(factor(length))
         lengthc = contr.sum(factor(length))  # use sum coding
        )

@@

fit1_ri <- brm( ri ~ state_or_event + lengthc + (1|participantnum), family=bernoulli(), data = d1_stats1, chains = nchain)

fit1_cd <- brm( cd ~ state_or_event + lengthc + (1|participantnum), family=bernoulli(), data = d1_stats1, chains = nchain)

>>>>

# Experiment 1 data frame 
d1 <- read_csv(dpath_exp1) #%>%
d2 <- read_csv(dpath_exp2) #%>%

@@

```{r exp1stats, eval=FALSE, fig.height=4, fig.width=6}

d1_stats1 <- d1 %>% 
  mutate(ri = factor(case_when(response=="RC_IC" ~  NA_character_,
                        response=="RC" ~     "root",
                        response=="RD_ID" ~  NA_character_,
                        response=="RD" ~     "root",
                        response=="IC" ~     "immediate",
                        response=="ID" ~     "immediate",
                        TRUE ~ NA_character_ 
                        ), levels = c("immediate", "root")),
         cd = factor(case_when(response=="RC_IC" ~  "continuous",
                        response=="RC" ~     "continuous",
                        response=="RD_ID" ~  "delayed",
                        response=="RD" ~     "delayed",
                        response=="IC" ~     "continuous",
                        response=="ID" ~     "delayed",
                        TRUE ~ NA_character_ 
                        ), levels = c("delayed", "continuous")),
         length = factor(length, levels = c("short", "medium", "long")),
         state_or_event= factor(state_or_event, levels = c("event", "state")),
        )

d1_stats1ri <- d1_stats1 %>% 
  drop_na(ri)

d1_stats1cd <- d1_stats1 %>% 
  drop_na(cd)

# use sum coding 
#options(contrasts=c("contr.sum", "contr.poly"))
options(contrasts=c("contr.treatment", "contr.poly"))

@@

nchain <- 1 # for testing
niter <- 500 # for testing

nchain <- 4 # for real run
niter <- 2000 # for real run

fit1_ri <- brm( ri ~ state_or_event + length + (1|participantnum), family=bernoulli(), data = d1_stats1ri, chains = nchain, iter = niter )

fit1_cd <- brm( cd ~ state_or_event + length + (1|participantnum), family=bernoulli(), data = d1_stats1cd,  chains = nchain, iter = niter )

cd_tab <- fixef(fit1_cd) %>% as_tibble(rownames="NA")

mcdtableout<- xtable(cd_tab, caption = "Coefficients for cd regression, Expt 1" )
print(mcdtableout, file = here("analysis", "output", paste0("coeftable_cd_exp1.tex")))

ri_tab <- fixef(fit1_ri) %>% as_tibble(rownames="NA")

mritableout<- xtable(ri_tab, caption = "Coefficients for ri regression, Expt 1" )
print(mritableout, file = here("analysis", "output", paste0("coeftable_ri_exp1.tex")), include.rownames = FALSE)

### Fig 2
```{r exp2plot, include=TRUE, fig.height=4, fig.width=6}

d2plot <- d2 %>% 
  filter(stimulustype != "chain") %>% 
  group_by(order, response) %>% 
  summarize(count = n()) %>% 
  mutate(response = factor(response, levels = c("RC", "RD", "IC", "ID", "other")))

plot2 <- d2plot %>% 
  ggplot(aes(x = response, y=count)) +
  geom_bar(stat="identity") +
  facet_wrap(~order)
plot2

### Stats for Experiment 2

```{r exp2stats, eval=TRUE, fig.height=4, fig.width=6}

d2_stats1 <- d2 %>% 
  mutate(ri = factor(case_when(response=="RC" ~     "root",
                               response=="RD" ~     "root",
                               response=="IC" ~     "immediate",
                               response=="ID" ~     "immediate",
                               TRUE ~ NA_character_ 
                               ), levels = c("immediate", "root")),
         cd = factor(case_when(response=="RC" ~     "continuous",
                               response=="RD" ~     "delayed",
                               response=="IC" ~     "continuous",
                               response=="ID" ~     "delayed",
                               TRUE ~ NA_character_ 
                               ), levels = c("delayed", "continuous")),
         order = factor(order, levels = c("RC_RD_ID", "RD_RC_ID", "RD_ID_RC"))
        )

d2_stats1ri <- d2_stats1 %>% 
  drop_na(ri)

d2_stats1cd <- d2_stats1 %>% 
  drop_na(cd)

# use sum coding 
#options(contrasts=c("contr.sum", "contr.poly"))
options(contrasts=c("contr.treatment", "contr.poly"))

m0 <- glmer(ri ~ order + (1|participantnum), data = d2_stats1, 
            family = binomial(link = "logit"))

nchain <- 1 # for testing
niter <- 500 # for testing

nchain <- 4 # for real run
niter <- 2000 # for real run

fit2_ri <- brm( ri ~ order + (1|participantnum), family=bernoulli(), data = d2_stats1ri, chains = nchain, iter = niter )

fit2_cd <- brm( cd ~ order + (1|participantnum), family=bernoulli(), data = d2_stats1cd,  chains = nchain, iter = niter )

cd_tab <- fixef(fit2_cd) %>% as_tibble(rownames="NA")

mcdtableout<- xtable(cd_tab, caption = "Coefficients for cd regression, Expt 2" )
print(mcdtableout, file = here("analysis", "output", paste0("coeftable_cd_exp2.tex")))

ri_tab <- fixef(fit2_ri) %>% as_tibble(rownames="NA")

mritableout<- xtable(ri_tab, caption = "Coefficients for ri regression, Expt 2" )
print(mritableout, file = here("analysis", "output", paste0("coeftable_ri_exp2.tex")), include.rownames = FALSE)

### 2021-01-31 docs/resources/CogSci_paper/paper.tex
figures modified
aurelien edits

Psychological theories of actual causation aim to characterize which of multiple causes of an event is singled out as the primary cause. We present one such theory called the continuity account of actual causation. The continuity account treats events as changes of state in continuous time and traces a sequence of stage changes backwards through time from an event to its primary cause. The account is broadly compatible with the physical process view of causation, and we test it by asking people to identify the primary cause of events occurring in simple physical systems. An initial experiment confirms that root causes are more likely to be chosen as primary causes than are immediate causes. A second experiment demonstrates that root causes that have temporal continuity with the effect are preferred even when probability raising accounts would predict otherwise. The results of both experiments are consistent with the continuity account, and suggest that inferences about changes of state in continuous time may underpin an important class of actual causation judgments.

@@

The following sections introduce the continuity account of actual causation and present two experiments designed to test core predictions of the account. Among previous theories of actual causation the most natural comparison is Spellman's probability raising account, which proposes that the actual cause of an effect is the cause that increased its probability to the greatest extent~\cite{spellman97}. Our second experiment directly compares the continuity account with the probability raising account and we find that the continuity account provides the better account of our data.

@@

\begin{figure*}
\begin{center}
\null\hfill
\subfloat[\label{subfig:1a}]{%
\includegraphics{figures/fig1a.eps}
}
\hfill
\subfloat[\label{subfig:1b}]{%
\includegraphics{figures/fig1b.eps}
}
\hfill\null
\end{center}
\caption{(a)(i) A network of particle detectors. All components are initially inactive. (ii) Detector \ev{R}{D} activates and activity propagates along the top branch. (iii) Detector \ev{R}{C} activates and ultimately triggers the activation of the GCM, which activates only when both  \ev{I}{D} and \ev{I}{C} are active. (b) An activation sequence in which (i) the top branch activates first, (ii) activation starts and finishes along the bottom branch while activation continues along the top branch, and (iii) the top branch completes its activation and ultimately triggers the GCM.}
%\caption{(a)(i) A smoke alarm connected to a chain of ionization detectors (top branch) and a chain of photoelectric detectors (bottom branch). All components are initially inactive (ii) Detector \ev{R}{D} activates and activity propagates along the top branch. (iii) Detector \ev{R}{C} activates and ultimately triggers the activation of the alarm, which activates only when both  \ev{I}{D} and \ev{I}{C} are active. (b) An activation sequence in which (i) the top branch activates first, (ii) activation starts and finishes along the bottom branch while activation continues along the top branch, and (iii) the top branch completes its activation and ultimately triggers the smoke alarm.}
\label{fig:intro}
\end{figure*}

@@

Because both of these general accounts of causation make no clear predictions, it seemed possible that people's inferences about the networks in our experiments would be highly variable and would reveal no clear trends. Our first experiment therefore explored simple cases analogous to the smoke detector example in Figure~\ref{fig:intro} with the goal of establishing whether the basic experimental procedure was viable.  The second experiment focused on more elaborate dual-branch cases that aimed to distinguish between the continuity account and Spellman's probability raising account~\cite{spellman97}.

@@

\textbf{Design}.  The experiment included activation sequences over 15 networks (3 chains and 12 dual branch networks). Excluding the GCM, each chain and each branch of each dual network had 7 detectors. We refer to the sequences as \emph{short}, \emph{medium} or \emph{long} based on the distance between the root causes and the GCM. Excluding the GCM, the short, medium and long chain sequences showed 1, 3 and 6 active detectors respectively at the end of the sequence. Within each dual branch sequence, the root causes on the two branches (\ev{R}{D} and \ev{R}{C}) were equidistant from the GCM, but these distances varied across sequences. The sequence in Figure~\ref{fig:intro}a is a medium dual branch sequence (3 active detectors in each branch excluding the GCM), and short and long versions had 1 and 6 active detectors respectively per branch.

@@

\begin{figure}[t]
\begin{center}
%\includegraphics[width=8cm]{results_E2}
\includegraphics{figures/expt2.eps}
%\includegraphics{figures/expt2test.eps}
%\includegraphics{figures/results_E2.eps}
\end{center}
\caption{Experiment 2. (a) The probability raising and continuity accounts make different predictions for dual branch sequences in which \ev{R}{C} occurs before \ev{R}{D}. (b) The two accounts agree when \ev{R}{C} occurs after \ev{R}{D}.
(c) Responses for the three possible orders of the key events.} 
\label{fig:expt2}
\end{figure}

@@

The continuity account highlights ideas such as the ``no coincidences'' principle that go beyond default formulations of the counterfactual and physical process accounts of causation. Key predictions of the account, however, can be reconstructed within both of these general frameworks. The counterfactual account can exploit fine-grained temporal information if each event is supplemented with a timestamp. If the effect to be explained in Figure~\ref{fig:intro} ``the activation of the GCM at exactly 10 sec after the start of the trial,'' then the counterfactual account suggests that \ev{R}{C} is a cause of the effect but \ev{R}{D} is not, because if \ev{R}{D} had been different (e.g.\ by occurring slightly earlier or later) then the effect would still have occurred at the same instant.

As suggested earlier, standard formulations of the physical process view treat both \ev{R}{C} and \ev{R}{D} as causes of the effect in Figure~\ref{fig:intro}. Our data are consistent with the possibility that people view \ev{R}{C} and \ev{R}{D} as equally valid causes, but pick \ev{R}{D} when forced to break the tie for reasons that may be relatively superficial. For example, perhaps \ev{R}{D} is preferred because the activation on the continuous branch is continuous both in space and time, and therefore more prototypical of a causal process than the activation along the delay branch. We have planned a future study that asks whether the \ev{R}{C} over \ev{R}{D} is fundamental or superficial using a task in which participants rank multiple candidate causes in order of importance, and in which ties are permitted.

>>>>

Psychological accounts of actual causation aim to characterize which of multiple causes of an event is singled out as the primary cause. We present one such account called the continuity account. The continuity account treats events as changes of state in continuous time and traces a sequence of stage changes backwards through time from an event to its primary cause. The account is broadly compatible with the physical process view of causation, and we test it by asking people to identify the primary cause of events occurring in simple physical systems. An initial experiment confirms that root causes are more likely to be chosen as primary causes than are immediate causes. A second experiment demonstrates that root causes that have temporal continuity with the effect are preferred even when probability raising accounts would predict otherwise. The results of both experiments are consistent with the continuity account, and suggest that inferences about changes of state in continuous time may underpin an important class of actual causation judgments.

@@

The following sections introduce the continuity account of actual causation and present two experiments designed to test core predictions of the account. Among previous theories of actual causation a natural comparison is Spellman's probability raising account, which proposes that the actual cause of an effect is the cause that increased its probability to the greatest extent~\cite{spellman97}. Our second experiment directly compares the continuity account with the probability raising account and we find that the continuity account provides the better account of our data.

@@

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{center}
\null\hfill
\subfloat[]{\includegraphics{figures/fig1a.eps}\label{subfig:1a}}
\hfill
\subfloat[]{\includegraphics{figures/fig1b.eps}\label{subfig:1b}}
\hfill\null
\end{center}
\caption{(a)(i) A network of particle detectors. All components are initially inactive. (ii) Detector \ev{R}{D} activates and activity propagates along the top branch. (iii) Detector \ev{R}{C} activates and ultimately triggers the activation of the GCM, which activates only when both  \ev{I}{D} and \ev{I}{C} are active. (b) An activation sequence in which (i) the top branch activates first, (ii) activation starts and finishes along the bottom branch while activation continues along the top branch, and (iii) the top branch completes its activation and ultimately triggers the GCM.}
%\caption{(a)(i) A smoke alarm connected to a chain of ionization detectors (top branch) and a chain of photoelectric detectors (bottom branch). All components are initially inactive (ii) Detector \ev{R}{D} activates and activity propagates along the top branch. (iii) Detector \ev{R}{C} activates and ultimately triggers the activation of the alarm, which activates only when both  \ev{I}{D} and \ev{I}{C} are active. (b) An activation sequence in which (i) the top branch activates first, (ii) activation starts and finishes along the bottom branch while activation continues along the top branch, and (iii) the top branch completes its activation and ultimately triggers the smoke alarm.}
\label{fig:intro}
\end{figure*}

@@

Because both of these general accounts of causation make no clear predictions, it seemed possible that people's inferences about the networks in our experiments would be highly variable and would reveal no clear trends. Our first experiment therefore explored simple cases analogous to the example in Figure~\ref{fig:intro}a with the goal of establishing whether the basic experimental procedure was viable.  The second experiment focused on more elaborate dual-branch cases that aimed to distinguish between the continuity account and Spellman's probability raising account~\cite{spellman97}.

@@

\textbf{Design}.  The experiment included activation sequences over 15 networks (3 chains and 12 dual branch networks). Excluding the GCM, each chain and each branch of each dual network had 7 detectors.  Within each dual branch sequence, the root causes on the two branches (\ev{R}{D} and \ev{R}{C}) were equidistant from the GCM, but these distances varied across sequences.
We refer to the sequences as \emph{short}, \emph{medium} or \emph{long} based on the distance between the root causes and the GCM. Excluding the GCM, the short, medium and long sequences showed 1, 3 and 6 active detectors respectively per branch at the end of the sequence.  The sequence in Figure~\ref{fig:intro}a is a medium dual branch sequence (3 active detectors in each branch excluding the GCM).

@@

\begin{figure}[t]
\begin{center}
%\includegraphics[width=8cm]{results_E2}
%\includegraphics{figures/expt2.eps}
\includegraphics{figures/expt2test.eps}
%\includegraphics{figures/results_E2.eps}
\end{center}
\caption{Experiment 2. (a) The probability raising and continuity accounts make different predictions for dual branch sequences in which \ev{R}{C} occurs before \ev{R}{D}. (b) The two accounts agree when \ev{R}{C} occurs after \ev{R}{D}.
(c) Responses for the three possible orders of the key events.} 
\label{fig:expt2}
\end{figure}

@@

The continuity account highlights ideas such as the ``no coincidences'' principle that go beyond default formulations of the counterfactual and physical process accounts of causation. Key predictions of the account, however, can be reconstructed within both of these general frameworks. The counterfactual account can exploit fine-grained temporal information if each event is supplemented with a timestamp. If the effect to be explained in Figure~\ref{fig:intro} is ``the activation of the GCM at exactly 10 sec after the start of the trial,'' then the counterfactual account suggests that \ev{R}{C} is a cause of the effect but \ev{R}{D} is not, because if \ev{R}{D} had been different (e.g.\ by occurring slightly earlier or later) then the effect would still have occurred at the same instant.

As suggested earlier, standard formulations of the physical process view treat both \ev{R}{C} and \ev{R}{D} as causes of the effect in Figure~\ref{fig:intro}. Our data are consistent with the possibility that people view \ev{R}{C} and \ev{R}{D} as equally valid causes, but pick \ev{R}{C} when forced to break the tie for reasons that may be relatively superficial. For example, perhaps \ev{R}{C} is preferred because the activation on the continuous branch is continuous both in space and time, and therefore more prototypical of a causal process than the activation along the delay branch. We have planned a future study that asks whether the \ev{R}{C} over \ev{R}{D} is fundamental or superficial using a task in which participants rank multiple candidate causes in order of importance, and in which ties are permitted.
